The Efficient Handling of Guards in the Design of RPython's Tracing JIT

### 2012-08-15 talk/vmil2012/paper.tex
attempt at a different introduction

\todo{the introduction needs some work}
\cfbolz{the first two two paragraphs talk about deoptimization, then it
switches to guards. I would say we should only talk about guards in the
beginning}
Tracing just-in-time (JIT) compilers record and compile linear control flow paths of operations executed by an interpreter
inserting operations called guards at points of possible divergence.
In this paper we describe and analyze how guards work, explaining what concepts
are used in the intermediate and low-level representation of the JIT
instructions and how these are implemented.

Our aim is to help understand the design constraints when
implementing guards. Guards have a runtime cost, they take time to execute. On
the other hand, guards are possible deoptimization points, meaning the recorded
and compiled path has to be left returning control to the interpreter. They need
enough associated information to enable rebuilding the interpreter state.
Based on the informal observation that guards are among the most common
operations in the traces produced by RPython's tracing JIT, our
goal is to present concrete numbers for the frequency and the overhead related
to guards, explain how they are implemented in the different levels of this particular
tracing JIT and clarify the rationale behind the design decisions based on the
numbers provided here.

The operations executed by an interpreter are recorded by the tracing JIT in
case they are frequently executed (this process is described in more detail in
Section~\ref{sec:Resume Data}). During the recording phase guards are
inserted into the recorded trace at all
points where the control flow could diverge. As can be seen in
Figure~\ref{fig:guard_percent} guards account for about 14\% to 22\% of the
operations before and for about 15\% to 20\% of the operations after
optimizing the traces generated for the different benchmarks used in this paper.
This makes guards one of the most common types of operations. As this paper will show, many of these guards
fail rarely or not all during execution.

There are several aspects to consider
in the design and optimization of guards, the first aspect is that due to the
large number of guards the memory overhead related to storing the information
needed for deoptimization should be kept low. A second aspect is that
successfully checking guards, i.e. not leaving the compiled trace, – which is
the common case – should be a cheap operation to execute favouring the on-trace
execution speed in contrast to the deoptimization case where the state has to
be rebuilt using the stored information. These constraints and trade-offs are
what make the design and optimization of guards an important and non-trivial
aspect of the low-level design of a tracing just-in-time compiler.

>>>>

\todo{the introduction needs some work}
\cfbolz{the first two two paragraphs talk about deoptimization, then it
switches to guards. I would say we should only talk about guards in the
beginning}

Tracing just-in-time (JIT) compilers record and compile commonly executed
linear control flow paths consisting of operations executed by an interpreter.
At points of possible divergence from the traced path operations called guards
are inserted. Furthermore, type guards are inserted to specialize the trace
based on the types observed during tracing. In this paper we describe and
analyze how guards work and explain the concepts used in the intermediate and
low-level representation of the JIT instructions and how these are implemented.
This is done in the context of the RPython language and the PyPy project, which
provides a tracing JIT compiler geared at dynamic language optimization.

Our aim is to help understand the design constraints when implementing guards
and to describe the concrete techniques used in the various layers of RPython's
tracing JIT. All design decisions will be motivated by concrete numbers for the
frequency and the overhead related to guards.

It is important to handle guards well, because they are very common operations
in the traces produced by tracing JITs. As can be seen in
Figure~\ref{fig:guard_percent} guards account for about 14\% to 22\% of the
operations before and for about 15\% to 20\% of the operations after optimizing
the traces generated for the different benchmarks used in this paper. An
additional property is that guard failure rates are very uneven. The majority
of guards never fail at all, whereas those that do usually fail extremely
often.

Besides being common, guards have various costs attached to them.
Guards have a runtime cost, they take time to execute. Therefore it is
important to make the on-trace execution of guards as efficient as possible. On
the other hand, guards are possible deoptimization points. The recorded and
compiled path has to be left if a guard fails, returning control to the
interpreter. Therefore guards need enough associated information to enable
rebuilding the interpreter state. The memory overhead of this information
should be kept low. These constraints and trade-offs are what make the design
and optimization of guards an important and non-trivial aspect of the low-level
design of a tracing just-in-time compiler.
